---
title: "geo_data"
author: "Zach Stone"
date: "2022-08-05"
output: html_document
---

```{r}
library(readxl)
library(tidyr)
library(dplyr)
library(ggplot2)
```

# Clean census data
```{r}
xl_data = read_excel('data/nyc_decennialcensusdata_2010_2020_change.xlsx', sheet = "2010, 2020, and Change")
```


```{r}
#The header is like a multiindex concatenated across the sections of the table
header = xl_data[1:3,]
header
```

```{r}
# On the geography section, it is just the second level which contains column names
header[,1:8]
```

```{r}
#Get column names from second level of header
names(xl_data) = as.character(xl_data[3,])
xl_data

#Slice data
xl_data = xl_data[4:nrow(xl_data),]
xl_data = xl_data %>% rename(index = `Orig Order`) %>% mutate(index = as.numeric(index))
```

```{r}
#The data can be broken into analysis from different surveys
xl_data %>%
  select(GeoType) %>%
  table()
```

```{r}
#Filter the census data
census = xl_data %>%
  filter(GeoType == 'CT2020')

#Name, CD Type, NTA Type are not used in this data
colSums(is.na(census)) == nrow(census)

census = census %>%
  select(-(6:8))
```

GeoID is 11 characters, which means it is a census tract ID. Split into:
-     state_gid (2 digit state GeoID)
-     county_gid (3 digit county GeoID)
-     tract_gid (6 digit census tract GeoID)

```{r}
census = census %>%
  mutate(StateGID = substring(GeoID, 1,2),
         CountyGID = substring(GeoID, 3,5),
         TractGID = substring(GeoID, 6,11))
```

Clearly the StateGID is unique

```{r}
unique(census$StateGID)
```

And counties correspond to Borough

```{r}
census %>%
  group_by(Borough, CountyGID) %>%
  summarize(tracts = n())
```

In all cases, redundant tract gid's are in distinct burroughs. 
So, sufficient to keep  Borough and tract_id from the Geography section

```{r}
has_dup = census %>%
  select(TractGID) %>%
  table() > 1

duplicates <- names(has_dup[has_dup])

census %>%
  filter(TractGID == duplicates[400])
```
```{r}
census = census %>%
  select(-c('GeoType', 'BCT2020', 'StateGID', 'CountyGID'))
```

Split into dataframes for 2010, 2020, and Change
```{r}
census2010 = census %>%
  select(index, Borough, TractGID, GeoID, ends_with('_10'))

census2020 = census %>%
  select(index, Borough, TractGID, GeoID, ends_with('_20'))

census_ch = census %>%
  select(index, Borough, TractGID, GeoID, ends_with('_Ch'))
```


```{r}
#Convert to numeric
census2020[,4:ncol(census2020)] = data.frame(sapply(census2020[,4:ncol(census2020)], as.numeric))
census2020$TractGID = as.numeric(census2020$TractGID)
```

```{r}
census2020
```

# Import GeoJSON file

```{r}
library(rjson)
library(geojsonio)
library(sp)
```


```{r}
spdf <- geojson_read('data/census2020.json', what = 'sp')
```

```{r}

plot(spdf)
```

```{r}
pop_table = census2020 %>% select(GeoID, Pop_20)
pop_table
spdf@data
```


```{r}
spdf_pop <- merge(spdf, pop_table, by.x = "GEOID", by.y = "GeoID")
```

```{r}
#Successfully added Pop_20 column
spdf_pop@data

length(spdf_pop)
length(spdf)
```

```{r}
pal = hcl.colors(200, palette = 'Purple-Orange')
max_pop = max(spdf_pop@data$Pop_20)
get_color = function (size) {
  index = round(199 * (size / max_pop) + 1, digits = 1)
  return(pal[index])
}

pop_col = sapply(spdf_pop$Pop_20, get_color)
```

```{r}
plot(spdf_pop, col = pop_col)
legend(x = -70, legend = 'Population', col = pop_col)
```
Using spplot

```{r}
spplot(spdf_pop, zcol = 'Pop_20')
```

```{r}
census2020
```


# ZIP codes

Import census tract to zip conversion

```{r}
tract_zip = read_excel('data/TRACT_ZIP_122020.xlsx')
tract_zip = tract_zip %>%
  select(tract = TRACT, zip = ZIP) %>%
  mutate(tract = as.numeric(tract), zip = as.numeric(zip))
```

Create conversion table

```{r}
#Conversion is not unique

# tract_zip %>%
#   group_by(tract) %>%
#   summarize(n())
```


```{r}
#Check that random selection is working as expected
# set.seed(19)
# tract_zip %>%
#   group_by(tract) %>%
#   summarize(z1 = nth(zip,sample(1:n(),size = 1)), z2 = nth(zip,sample(1:n(),size = 1)), n()) 
```

```{r}
#randomly select zip code for each ambiguous tract number
set.seed(17)
zip_table = tract_zip %>%
  group_by(tract) %>%
  summarize(zip = nth(zip, sample(1:n(),size = 1)))
```



```{r}
#still has many missing zip codes due to imperfect location encoding by zip code
census2020 %>%
  left_join(zip_table, by = c("GeoID" = "tract")) %>%
  select(c(1:4,ncol(census2020)+1,5:ncol(census2020))) 

census2020
```

## Testing plotting aggregate geographic data

```{r}
census2020
spdf_zip@data
zip_table
```


```{r}
library(rgeos)
```

```{r}
sp_zip = gUnaryUnion(spdf_zip, id = spdf_zip$zip)
```

```{r}
pop_zip = census2020 %>%
  inner_join(zip_table, by = c('GeoID' = 'tract')) %>%
  group_by(zip) %>%
  summarize(pop = sum(Pop_20))

row.names(pop_zip) = c(pop_zip$zip)

spdf_zipped = SpatialPolygonsDataFrame(sp_zip, pop_zip, proj4string = CRS('+proj=longlat +datum=WGS84 +no_defs'))
```



```{r}
spplot(spdf_zipped, zcol = 'pop')
```
### Final check using area

```{r}
zip_area = spdf_zip@data %>%
  group_by(zip) %>%
  summarize(area = sum(Shape__Area))

row.names(zip_area) = zip_area$zip

spdf_ziparea = SpatialPolygonsDataFrame(sp_zip, zip_area)
```

```{r}
#Sure looks like it plotted sums of census tract areas
spplot(spdf_ziparea, zcol = 'area')
```

## Checking adding points and matching coordinates

```{r}
crs(spdf)
```


```{r}
cents = getSpPPolygonsLabptSlots(spdf_zipped)
spdf_cent = SpatialPointsDataFrame(coords=cents, data=spdf_zipped@data, proj4string=CRS('+proj=longlat +datum=WGS84 +no_defs'))
```

```{r}
plot(spdf_zipped)
points(spdf_cent, pch = 3, col = "Red")
```

```{r}
plot(spdf_cent)
plot(spdf_zipped)
```

```{r}
gArea(spgeom = spdf_zipped)
```
 
 
```{r}
SpatialPointsDataFrame(coords=cents, data=sids@data, proj4string=CRS("+proj=longlat +ellps=clrk66"))
```




# Medical data (by conditions)

```{r}
er_data = read.csv('data/Medicaid_Chronic_Conditions__Inpatient_Admissions_and_Emergency_Room_Visits_by_Zip_Code__Beginning_2012.csv')
```

```{r}
er = er_data
str(er)
```

```{r}
cty_to_boro <- function(county) {
  switch(county, 'Richmond' = 'Staten Island', 
         'Kings' = 'Brooklyn', 'New York' = 'Manhattan', 
         'Queens' = 'Queens', 'Bronx' = 'Bronx')
}

ernyc = er %>%
  filter(Primary.County %in% c('Kings', 'New York','Richmond', 'Queens', 'Bronx')) %>%
  mutate(Primary.County = sapply(Primary.County, cty_to_boro)) %>%
  rename(Boro = Primary.County) %>%
  arrange(Year, Boro)
```

```{r}
ernyc
```
```{r}
#Just testing joining data, though will mostly likely want to work within the dataframe and zip code-encoded census data
# census2020 %>%
#   inner_join(zip_table, by = c("GeoID" = "tract")) %>%
#   select(Borough, GeoID, zip, Pop_20) %>%
#   group_by(Borough,zip) %>%
#   summarize(total_pop = sum(Pop_20)) %>% ungroup() %>%
#   inner_join(ernyc, by = c("zip" = "Zip.Code")) %>%
#   select(4:5,2,3,6:ncol(.))
```

```{r}
ggplot(ernyc, aes(y = Major.Diagnostic.Category, fill = Boro)) + geom_bar() 
```

```{r}
ernyc %>%
  filter(Episode.Disease.Category == 'Asthma') %>%
  ggplot(data = ., aes(x = Boro)) + geom_bar()
```
```{r}
ernyc
```




# Medical data (region totals)

```{r}
ertotals_data = read.csv('data/Medicaid_Beneficiaries__Inpatient_Admissions_and_Emergency_Room_Visits_by_Zip_Code__Beginning_2012.csv')
```

```{r}
ertotals = ertotals_data
str(ertotals)
```

```{r}
ertotals %>%
  filter(County %in% c('Kings', 'New York','Richmond', 'Queens', 'Bronx')) %>%
  mutate(County = sapply(County, cty_to_boro)) %>%
  rename(Boro = County) %>%
  arrange(Year, Boro)
```

# Tobacco dataset inspect

```{r}
tobac_data = read.csv('data/Active_Tobacco_Retail_Dealer_Licenses.csv')
tobac = tobac_data
```

```{r}
#All License.Types are business
tobac %>%
  group_by(License.Type) %>%
  summarize(n())

#All in same industry
tobac %>%
  filter(License.Status == 'Active') %>%
  group_by(Industry) %>%
  summarize(n())

tobac %>%
  filter(License.Status == 'Active') %>%
  select(id = DCA.License.Number, city = Address.City, zip = Address.ZIP,
         boro = Address.Borough, tractGID = Census.Tract, long = Longitude, lat = Latitude)
```


# Income

```{r}
income_data = read.csv('data/census_income.csv')
income = income_data
names(income) = c(income[1,])
income = income[-1,]
```


```{r, warning=FALSE}
source('census_import.R')
income = income %>%
  mutate(id = gsub(pattern = '1400000US', replacement = '', x = id)) %>%
  census_num(omit_cols = 'Geographic Area Name') %>%
  rename(GeoID = id)
```

```{r}
header
```

```{r}
income2020 = census2020 %>%
  transmute(GeoID, population = Pop_20, housed_pop = HHPop_20, 
            households = HHPop_20/AvHHSz_20) %>%
  inner_join(income %>% select(GeoID, contains('!!Households!!Mean')),
             by = 'GeoID') %>%
  select(-6)

names(income2020) = c('GeoID', 'population', 'housed_pop', 'households', 'mean_HH_income')
```

```{r}
income2020 %>%
  transmute(GeoID, housed_pop, households, mean_HH_income, net_HH_income = households * mean_HH_income)
```

```{r}

```


