---
title: "R Notebook"
output: html_notebook


---
# Set up and imports
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/Users/zachstone/Documents/coursework/NYCDSA/R_project')
```

```{r}
library(sp)
library(raster)
```

```{r}
library(rjson)
library(geojsonio)
```


```{r}
spdf <- geojson_read('./data/census2020.json', what = 'sp')
```

# Example SPDF
```{r}
plot(spdf)
```

```{r}
spdf@data
```
# Constructing SPDF from scratch

## Create raw polygon objects

```{r}
Sr1 = Polygon(cbind(c(2,4,4,1,2),c(2,3,5,4,2)), hole = TRUE)
Sr2 = Polygon(cbind(c(5,4,2,5),c(2,3,2,2)))
Sr3 = Polygon(cbind(c(4,4,5,10,4),c(5,3,2,5,5)))
Sr4 = Polygon(cbind(c(5,6,6,5,5),c(4,4,3,3,4)))
```

## Construct regions (Polygons objects with IDs)

```{r}
Srs1 = Polygons(list(Sr1), "s1") #made of single polygon Sr1
Srs2 = Polygons(list(Sr2), "s2") #ditto Sr2
Srs3 = Polygons(list(Sr3, Sr4), "s3/4") #region composed of  polygons, one 2d region the other 1d cycle.
Srs4 = Polygons(list(Sr4, Sr3), 's4/3')
Srs5 = Polygons(list(Sr1,Sr2), 's1/2')
Srs6 = Polygons(list(Sr1,Sr4), 's1/4')
```

```{r}
plot(SpatialPolygons(list(Polygons(list(Sr3), 'X'))), col = 1)
plot(SpatialPolygons(list(Polygons(list(Sr4), 'X'))), col = 1)
plot(SpatialPolygons(list(Srs3)), col = 2)
plot(SpatialPolygons(list(Srs4)), col = 2)
plot(SpatialPolygons(list(Srs5)), col = 2)
plot(SpatialPolygons(list(Srs6)), col = 2)
plot(SpatialPolygons(list(Srs1)), col = 2)
```

### Figuring out Polygons region construction. It looks like the region constructed by Polygons is the symmetric difference of the polygon objects in the list. Test with overlapping sets

```{r}
x1 = c(1,-1,-3,-1,1)
y1 = c(0,2,0,-2,0)

s1 = Polygon(cbind(x1,y1))
```

```{r}
x2 = c(3,1,-1,1, 3)
y2 = c(0,2,0, -2, 0)

s2 = Polygon(cbind(x2,y2))
```

```{r}
x3 = c(2,0,-2,0, 2)
y3 = c(1,3,1, -1, 1)

s3 = Polygon(cbind(x3,y3))
```

```{r}
#Unexpected behavior
#Official documentation says that regions should either be:
# - Disjoint, in which case it will union them, or
# - One contained fully in the other, in which case it will subtract the smaller
r = Polygons(list(s1,s2,s1), 'R')
plot(SpatialPolygons(list(r)), col = 1)
```

What about unions of regions inside?

```{r}
x1 = c(2,0,-2,0, 2)
y1 = c(0, 2, 0, -2, 0)

s1 = Polygon(cbind(x1,y1))
```

```{r}
x2 = c(1,1,-1,-1, 1)
y2 = c(0, 1, 1, 0, 0)

s2 = Polygon(cbind(x2,y2), hole = TRUE)
```

```{r}
#Even making this a hole does not work as expected
x3 = c(.5,.5,-.5,-.5, .5)
y3 = c(.25, .75, .75, .25, .25)

s3 = Polygon(cbind(x3,y3), hole = FALSE)
```

```{r}
x4 = c(.5,.5,-.5,-.5, .5)
y4 = -c(.25, .75, .75, .25, .25)

s4 = Polygon(cbind(x4,y4), hole = TRUE)
```

```{r}
r = Polygons(list(s1,s2,s4), 'R')
plot(SpatialPolygons(list(r)), col = 1)
```



## Create SpatialPolygons region family

```{r}
#SpatialPolygons family with 1,2,3 ordering
SpP = SpatialPolygons(list(Srs1,Srs2,Srs3), 1:3)
plot(SpP, col = 1:3, pbg="white")
```

## Creating SpDf

Create a dataframe whose rownames match those of the SpatialPolygons object

```{r}
df = data.frame(x=c(10,50,55), y=c('a','b','c'), z=c('x', 'y', 'z'), row.names=row.names(SpP))
```

Simply pass the SP object first, followed by the df.
```{r}
spdf_ = SpatialPolygonsDataFrame(SpP, data = df)
```

The df can then be plotted, using the col argument to specify a vector
Conveniently, we can grab a column, and use the values to select colors from a hue range which we can construct using rainbows, hcl.colors, or similar. 

We can then transform data into a heatmap by constructing a palette on the appropriate scale and assigning values to places in the scale. To change gradation, make the palette larger or smaller, or change the mapping from the values to the indices.

```{r}
plot(spdf_, col = hcl.colors(100, palette = 'Rocket')[spdf_$x + 10])
```

```{r}
hcl.colors(100, palette = 'Rocket')[3]
```

# Practicing with the NYC dataframe

```{r}
spdf@data
```
# Merging regions

```{r}
library(rgeos)
```

```{r}
tract_zip = read_excel('data/TRACT_ZIP_122020.xlsx')
tract_zip = tract_zip %>%
  select(tract = TRACT, zip = ZIP) %>%
  mutate(tract = as.numeric(tract), zip = as.numeric(zip))

set.seed(17)
zip_table = tract_zip %>%
  group_by(tract) %>%
  summarize(zip = nth(zip, sample(1:n(),size = 1)))
```

```{r}
spdf@data
zip_table
```

```{r}
#Only saves tracts with an associated zip code, and attaches the zip codes
spdf_zip = merge(spdf, zip_table, by.x = 'GEOID', by.y = 'tract', all = FALSE)
```

```{r}
# plot(spdf_zip)
```

```{r}
spdf_u = gUnaryUnion(spdf_zip, id = spdf_zip$zip)
```


```{r}
spdf
```

```{r}
plot(spdf_u)
```

```{r}
cents = getSpPPolygonsLabptSlots(spdf_u)

SpatialPointsDataFrame(coords=cents, data=sp@data, proj4string=CRS("+proj=longlat +ellps=clrk66"))
```

```{r}
#Row names are the grouped values. Nice :)
# row.names(spdf_u)
```

